### 2024-01-26 parts/starfree.tex
writing to publish
preparing the witdrawal

%! TeX program = xelatex
%! lang = en-US
\section{Beyond Polynomials}
\label{sec:deciding}

Let us now move on from polynomials to general \kl{$\Rel$-polyregular
functions}. As we will prove, \kl{commutativity} reduces $\ZPoly$ to a
combination of a finite number of polynomials, arranged using modulo counting.
We will first decide $\NPoly$ inside $\ZPoly$, and then leverage this to decide
$\NSF$ inside $\NPoly$, incidentally proving that $\ZSF \cap \NPoly = \NSF$.

@@

The goal of this section is to go beyond polynomials and characterize all
$\Rel$-polyregular functions that are $\Nat$-polyregular, under the assumption
that they are \kl{commutative}.

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    $\alpha_0, \dots, \alpha_k \in \Sigma^*$,
    for all $u_1, \dots, u_k \in \Sigma^*$,
    there exists a \kl{$\Nat$-rational polynomial} $P$
    satisfying for all $X_1, \dots, X_k \geq \omega$:
    \begin{equation*}
        f
        \left(
            \alpha_0 \prod_{i = 1}^k u_i^{\omega X_i} \alpha_i
        \right)
        = 
        P(X_1, \dots, X_k) \quad .
    \end{equation*}
\end{definition}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and one can effectively compute a representation of $f$.
\end{theorem}

To prove the theorem, let us first provide a nice characterization of
\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
commutative case.

\AP Given $\omega \in \Nat$, let us write $\phi \colon \Nat \to \set{0, \dots,
\omega} \times \set{0, \dots, \omega - 1}$ defined as $\phi(n) = (t, r)$ where
$t = \min(n, \omega)$, and $r$ is the rest of the Euclidian division of $n$ by
$\omega$. We let $\phi^\dagger \colon \Nat^k \to$.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function}. There exists a computable
    $\omega \in \Nat$,
    and computable 
    polynomials $\seqof{P_{\rho}}{\rho \in \Delta_\omega}$,
    such that for all $w \in \Sigma^*$,
    \begin{equation*}
        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) = P_{\phi(X_1, \dots, X_n)} (X_1 / \omega, \dots, X_n / \omega)
        \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    equivalent partitions of the word, but commutativity?.q
    Lemma 5.37 phd gaëtan.
\end{proof}

As a consequence, we conclude 
\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
    We use \cref{decompose-polynomial:lem}
    to effectively compute
    a polynomial representation of $f$. Remark that
    $f$ is a \kl{$\Nat$-polyregular function} if and only if
    all these polynomials are \kl{$\Nat$-rational polynomials},
    which is decidable thanks to
    \cref{decide-correct-poly:lem,corrected-version:thm}.
\end{proof}

@@

Note that this proves in the commutative case a conjecture that was 
formulated by Thesis Gaetan, and that is non trivial, because

\begin{conjecture}
    \label{nsf-zsf:conj}
    In the general case, the following equation holds for all $k \in \Nat$:
    \begin{equation*}
        \ZSF[k] \cap \NPoly
        = 
        \NSF[k]
        \quad .
    \end{equation*}
\end{conjecture}

As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
decidable inside $\NPoly$ under the assumption of commutativity.

\textbf{TODO: add the growth rate here, we are actually minimising it}.
\begin{corollary}
    One can decide if a \kl{commutative} \kl{$\Rel$-polyregular function}
    can be realized by a \kl{star-free $\Nat$-polyregular function},
    in which case the conversion is effective.
\end{corollary}
\begin{proof}
    From \cite[Theorem V.13]{LOPEZ23b}, it is decidable whether
    $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
    it is decidable whether $f \in \NPoly$.
    Furthermore, one can enumerate functions in $\NSF[k]$,
    and check equivalence to effectively represent $f$ as a
    \kl{star-free $\Nat$-polyregular function}.
\end{proof}

@@

\begin{example}
    A language $L$ is aperiodic if and only if 
    the indicator function is ultimately polynomial.
\end{example}
\begin{proof}
    TODO.
\end{proof}

@@

    \textbf{Induction.}
    Because $f$ is \kl{ultimately polynomial},
    there exists $N_0 \in \Nat$, 
    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
    such that
    \begin{equation*}
        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
        = 
        P(\seqof{X_a}{a \in \Sigma})
        \quad
        \text{ when }
        \forall a \in \Sigma, X_a \geq N_0
        \quad .
    \end{equation*}
    Furthermore, leveraging \cref{decompose-polynomial:lem},
    we obtain $\omega \in \Nat$
    and a family of polynomials 
    $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{0 \leq n_a < \omega}$
    such that.
    Because each of those polynomials coincide with $P$ over an infinite
    subgrid of $\Nat$,
    we conclude that all polynomials are in fact equal to a single one.
    we conclude that the polynomials 
    $P_{\seqof{n_a}{a \in \Sigma}}$ are all equal.

    Now, we conclude in particular that the function 
    computing $P$ is \kl{$\Nat$-polyregular}, because $f$ is,
    hence is computable by a \kl{star-free $\Nat$-polyregular function},
    because of \cref{nsf-polynomials:thm}.
    When fixing values of some indeterminate below $N_0$,
    we obtain a new function $g$, that remains commutative,
    remains ultimately polynomial, and remains \kl{$\Nat$-polyregular}.
    By induction hypothesis, this function $g$ is also 
    \kl{star-free $\Nat$-polyregular}, and we conclude
    that 
    $f = P + \sum g$ which remains star free.
\end{proof}

>>>>

%! TeX program = xelatex
%! lang = en-US
\section{Beyond Polynomials}
\label{beyond-polynomials:sec}

Let us now go beyond polynomials and consider the case of \kl{commutative}
\kl{$\Rel$-polyregular functions}. The hypothesis of \kl{commutativity} reduces
$\ZPoly$ to a combination of a finite number of polynomials, arranged by
counting modulo some numbers. This characterization will be proven in
\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
\cref{decompose-polynomial:lem} is actually a refinement of
\cref{n-poly-combinatorics:lem} under the extra assumption of
\kl{commutativity}.

\AP To formally state the lemma, we will need some notations that will
formalize the intuitive idea of counting tuples of integers modulo some
constant. Let $\omega \in \Nat$, we write the set of \intro{types modulo
$\omega$} $\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
    \dots, \omega - 1}$. To a number $n \in \Nat$, we associate its
    \intro{$\omega$-type} written $\moduloType[\omega](n)$ which is defined as the pair
    $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the rest of the Euclidian
    division of $n$ by $\omega$. This notion of type is lifted to vectors in
    $\Nat^p$ for any $p \in \Nat$ by pointwise application.

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $w \in \Sigma^*$,
    \begin{equation*}
        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) 
        = P_{\moduloType[\omega](X_1, \dots, X_n)}
        \left(
            \floor{X_1/\omega}, \dots, \floor{X_n/\omega}
        \right)
        \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    We apply
    \cref{n-poly-combinatorics:lem} to compute the $\omega$.
    Using this $\omega$,
    for any choice of
    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
    there exists a polynomial $P_r \in \Rel[X_1, \dots, X_p]$
    such that $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
        = P_r(X_1, \dots, X_p) \quad .
    \end{equation*}
    To get the full family of polynomials needed requires
    just to only iterate subsets of letters, and fix the other ones
    to be below the threshold $\omega^2$.
\end{proof}

@@

The goal of this section is to go beyond polynomials and characterize all
\kl{$\Rel$-polyregular functions} that are \kl{$\Nat$-polyregular}, under the
assumption that they are \kl{commutative}. We will provide a semantic property
in the form of \cref{k-combinatorial:def}, that is conjecture to characterize
$\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

\begin{definition}
    \label{k-combinatorial:def}
    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
    such that
    for all
    $\alpha_0, \dots, \alpha_k \in \Sigma^*$,
    for all $u_1, \dots, u_k \in \Sigma^*$,
    there exists a polynomial $P$
    such that $\translate{K}(P) \in \CorrectPoly$, and
    satisfying for all $X_1, \dots, X_k \geq \omega$:
    \begin{equation*}
        f
        \left(
            \alpha_0 \prod_{i = 1}^k u_i^{\omega X_i} \alpha_i
        \right)
        = 
        P(X_1, \dots, X_k) \quad .
    \end{equation*}
    We say that $f$ is \reintro{combinatorial}
    if it is \reintro{$(k,\Nat)$-combinatorial} for all $k \in \Nat$.
\end{definition}

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and one can effectively compute a representation of $f$.
\end{theorem}
\begin{proof}
    \textbf{TODO: we need to talk about the number of variables hehe}
    Let us prove that
    \cref{f-npoly-combi:item}
    implies \cref{f-combinatorial:item}. 
    We compute the $\omega$ by applying 
    \cref{n-poly-combinatorics:lem}, and then
    notice that the polynomial $P$ obtained is
    by definition a \kl{$\Nat$-rational polynomial}.

    For the converse implication, we use \cref{decompose-polynomial:lem} to
    effectively compute a polynomial representation of $f$. Using the
    assumption, all of these polynomials are actually \kl{$\Nat$-rational
    polynomials}. Let us now construct a new way to compute $f$, by first
    computing the \kl{$\omega$-type} of the input (which is possible in
    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
    which is possible in $\NPoly$.
\end{proof}

@@

As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
decidable inside $\NPoly$ under the assumption of commutativity,
and that the effective procedure can be used

\begin{corollary}
    One can decide if a \kl{commutative} \kl{$\Rel$-polyregular function}
    can be realized by a \kl{star-free $\Nat$-polyregular function},
    in which case the conversion is effective.
\end{corollary}
\begin{proof}
    From \cite[Theorem V.13]{LOPEZ23b}, it is decidable whether
    $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
    it is decidable whether $f \in \NPoly$.
    Furthermore, one can enumerate functions in $\NSF[k]$,
    and check equivalence to effectively represent $f$ as a
    \kl{star-free $\Nat$-polyregular function}.
\end{proof}

@@

\begin{example}
    A language $L$ is aperiodic if and only if 
    $\ind{L}$ is \kl{ultimately polynomial}.
\end{example}

@@

    \textbf{Induction.}
    Because $f$ is \kl{ultimately polynomial},
    there exists $N_0 \in \Nat$, 
    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
    such that
    \begin{equation*}
        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
        = 
        P(\seqof{X_a}{a \in \Sigma})
        \quad
        \text{ when }
        \forall a \in \Sigma, X_a \geq N_0
        \quad .
    \end{equation*}
    Furthermore, leveraging \cref{decompose-polynomial:lem},
    we obtain $\omega \in \Nat$
    and a family of polynomials 
    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^\Sigma}$
    that represent $f$.
    For all $t \in \ModuloTypes[\omega]^\Sigma$
    that have threshold greater than $\omega^2$,
    we conclude that $P_t$ and $P$ coincide (up to translation) over an infinite grid,
    hence we conclude that
    if all letters appear at least $N_1 \defined \max (\omega^2, N_0)$ times,
    then $f$ computes $P$. But this also shows that
    $\translate{N_1}(P)$ is a \kl{$\Nat$-rational polynomial}, hence is
    \kl{represented} by a function $f_1 \in \NSF[k]$.

    To compute $f$ on words where some letter appears fewer than
    $N_1$ times, we partially apply $f$ to obtain a function $g$
    over a restricted input alphabet. Using the induction hypothesis,
    $g$ belongs to $\NSF[k]$, and we can therefore (effectively) compute
    a representation of $f$ in $\NSF[k]$.
\end{proof}

### 2024-01-26 parts/transducers.tex
preparing the witdrawal

The goal of this section is to provide tools that do not require
\kl{commutative} \kl{$\Nat$-polyregular functions}. To that end, we introduce
the notion of \kl{$k$-residual transducer}, that generalizes the notion of
residual transducer introduced for \kl{$\Rel$-polyregular functions} in
\cite{LOPEZ23b}. We will first introduce the object, provide an algorithm (with
an oracle function) that computes said object, and finally, check that in
simple cases, the aperiodicity of a function can be read on its
\kl{$k$-residual transducer}. The main drawbacks of the generalization are the
following: one needs to decide whether a \kl{$\Rel$-polyregular function} is in
$\NPoly$, which is only known in the \kl{commutative} case, and the
characterization in terms of counter free also assumes \kl{commutativity}.
However, we believe that the introduction of this object is relevant.

The goal of the upcoming section is to provide a notion of residual transducer
for \kl{commutative} \kl{$\Nat$-polyregular functions}. The notion of
transducer is based on the \kl{marble transducers} \cite{EHB99}.

@@

Given the syntax of an \kl{$\mathcal{H}$ transducer}, let us now 
introduce its semantics, which is given by a simple recursion
scheme.
\begin{definition}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Nat$
    is defined inductively  as follows:
    \begin{itemize}
        \item $\aTransd(q, \varepsilon) \defined F(q)$;
        \item $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    \end{itemize}
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

\AP As in the case of regular automata, we generalize the function $\delta
\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
$\lambda(q,\varepsilon) \defined 0$, and $\lambda(q, au) \defined \app{\lambda(q,a)}{u}
+ \lambda(\delta(q,a), u)$. Following these equations, we obtain that
$\aTransd(q_0, uv) = \aTransd(\delta(q_0, u), v) + \lambda(q_0, u)(v)$.
Furthermore, let us notice that if $\mathcal{H}$ is closed under finite sums
and shifts, then $\lambda(q, u) \in \mathcal{H}$ for all $(q,u) \in Q \times
\Sigma^*$.

@@

@@

\begin{theorem}[Cite gaëtan]
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$,
        \item $f$ is computed by a \kl{$\NPoly[k-1]$-transducer}.
    \end{itemize}
    \begin{itemize}
        \item $f \in \NSF[k]$,
        \item $f$ is computed by a \kl{$\NSF[k-1]$-transducer}
            that is counter-free.
    \end{itemize}
\end{theorem}

\AP Let us now introduce the notion of \intro{residual} of a function $f
\colon \Sigma^* \to \Rel$, by defining $\app{f}{u} \defined w \mapsto f(uw)$.
The collection of \reintro{residuals} of a function $f$ is denoted $\Res(f)$
and is defined as the set of $\app{f}{u}$ where $u$ ranges over words in
$\Sigma^*$.

\AP Let $k \in \Nat$. The residual transducer in the case of $\Rel$ output was
built using an equivalence relation defined as $f \equiv_k g$ if and only if $g
- f \in \ZPoly[k-1]$. However, in the case of $\Nat$ output, the relation
cannot be symmetric, and furthermore, we are going to order words rather than
residuals. Namely, given $k \in \Nat$, and $f \colon \Sigma^* \to \Nat$, we
define $u \resleq{f}{k} v$ if and only if $\app{f}{v} - \app{f}{u} \in
\NPoly[k-1]$.

@@

There is one major issue arising when defining the notion of residual
transducer based on the quasi-ordering $\resleq{f}{k}$. The naïve approach is
to consider as states some representatives for the $\resleq{f}{k}$ minimal
elements of $\Sigma^*$, all computing the same function
(\cref{good-residual-ordering:fact}), and define transitions by letting
$\delta(u, a)$ be \emph{one} state $v$ such that $v \resleq{f}{k} ua$. This
naïve approach does not yield a canonical model.
For instance, the function
of \cref{non-canonical-transd:ex} can be computed in two ways if we only
consider $\resleq{f}{k}$ minimal elements, as witnessed in
\cref{non-canonical-transd:fig}. Even worse, the function $f$ belongs to
$\NSF[0]$, but one of the two presented transducers has a \kl{counter}.

This is why in \cref{residual-transducer:def}, we introduce the prefix ordering
$\prefleq$ to further constrain how the model computes.

@@

\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
quasi-ordered set $(X, \leq)$ is \kl{good} whenever there exist $i < j$ such
that $u_i \leq u_j$. The set $X$ is a \kl{well-quasi-ordering} when
every infinite sequence is \kl{good}. A sequence is \kl{bad}
when it is not \kl{good}.

\begin{lemma}
    \label{wqo-implies-termination:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and $k \in \Nat$ be such that
    every infinite, $\prefleq$-increasing sequence is \kl{good}
    in $(\Sigma^*, \resleq{f}{k})$.
    Then, \cref{residual:algo} terminates on the input $(f,k)$.
\end{lemma}

@@

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item $f \in \NPoly[k]$;
        \item $(\Sigma^*, \resleq{f}{k})$ is a
            \kl{well-quasi-ordering};
        \item every $\prefleq$-increasing sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k}$;
        \item $f$ is \kl{computed} by a
            \kl{$k$-residual transducer}.
    \end{enumerate}
    If $f$ is asumed to be \kl{commutative}, then all of the above
    properties are decidable, and the conversions effective.
\end{theorem}
\begin{proof}
    The first two items are known to be equivalent since 
    the phd of gaëtan.
    We have proven that transducers imply
    wqo, and that wqo implies $k$-residual transducer.
\end{proof}

>>>>

The goal of this section is to provide tools that do not require the assumption
of \kl{commutativity}. To that end, we introduce the notion of \kl{$k$-residual
transducer}, that generalizes the notion of residual transducer introduced for
\kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}. This notion of residual
transducer has its own interest, as it shifts the attention from (compatible)
equivalence relations of finite index that are the classical tool in automata
theory, towards (compatible) order relations that are \kl{well-quasi-ordered},
which is the order counterpart of having finite index.

While \kl{$k$-residual transducers} will provide canonical objects associated
to functions in $\NPoly$, their effective construction requires the ability to
decide $\NPoly$ inside $\ZPoly$, which is not currently known (except in the
commutative case). Furthermore, the correspondence between \kl{counters} in
this automaton and \kl{star-free $\Nat$-polyregular functions} is non-trivial,
and cannot be used \emph{as is} to decide membership in $\NSF$
based on this analysis of the \kl{$k$-residual transducer}.

\AP The upcoming sections will heavily refer to the notion of \intro{residual} of
a function $f \colon \Sigma^* \to \Rel$,  defined by $\app{f}{u} \defined w
\mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
denoted $\Res(f)$ and is defined as the set of $\app{f}{u}$ where $u$ ranges
over words in $\Sigma^*$.

\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
that $u_i \leq u_j$. The set $X$ is a \intro{well-quasi-ordering} when every
infinite sequence is \reintro{good}. A sequence is \intro{bad} when it is not
\reintro{good}. The notion of \reintro{good sequences} can also be applied to
binary relations that are not orderings, and a binary relation $R$ for which
every infinite sequence is \reintro{good} is said to be \intro{well}
\cite{MELL98}.

\subsection{Residual Transducers and WQOs}

The goal of the upcoming section is to provide a notion of residual transducer
for \kl{commutative} \kl{$\Nat$-polyregular functions}. The notion of
transducer is based on the marble transducers \cite{EHB99}, and follows the
general pattern of transducers introduced in \cite{LOPEZ23b}.

@@

Given the syntax of an \kl{$\mathcal{H}$-transducer}, let us now 
introduce its semantics, which is given by a simple recursion
scheme.
\begin{definition}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Nat$
    is defined inductively  as follows:
    \begin{itemize}
        \item $\aTransd(q, \varepsilon) \defined F(q)$;
        \item $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    \end{itemize}
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

@@

\AP As in the case of regular automata, we generalize the function $\delta
\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
$\lambda(q,\varepsilon) \defined 0$, and recursively define $\lambda(q, au)
\defined \app{\lambda(q,a)}{u} + \lambda(\delta(q,a), u)$. Following these
equations, we can write for all $q \in Q$, $u,v \in \Sigma^*$, $\aTransd(q, uv)
= \aTransd(\delta(q, u), v) + \lambda(q, u)(v)$. Finally, let us notice
that if $\mathcal{H}$ is closed under finite sums and shifts, then $\lambda(q,
u) \in \mathcal{H}$ for all $(q,u) \in Q \times \Sigma^*$.

@@

It is already known in the literature that this notion of transducers correctly
captures both $\NSF$ and $\NPoly$ in the sense that the following fact will
clarify.

\begin{fact}[Cite gaëtan]
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
        \item $f$ is computed by an \kl{$\NPoly[k-1]$-transducer}
            (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
            \kl{counter-free}).
    \end{itemize}
\end{fact}

\AP Let $k \in \Nat$. We define $f \zpolyequiv[k] g$ if and only if $(g - f)
\in \ZPoly[k-1]$, and $f \npolyleq[k] g$ if and only if $(g - f) \in
\NPoly[k-1]$. Given a function $f \colon \Sigma^* \to \Rel$, we are going to
study both $(\Res(f), \zpolyequiv[k])$, and $(\Res(f), \npolyleq[k])$. However,
to simplify notations, instead of writing $\app{f}{u} \npolyleq[k] \app{f}{v}$
when $u,v \in \Sigma^*$, we will use the convenient notation $u \resleq{f}{k}
v$, and directly consider the space $(\Sigma^*, \resleq{f}{k})$.

@@

Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
approach is to consider as states some representatives for the $\resleq{f}{k}$
minimal elements of $\Sigma^*$ (assuming there are finitely many of them), all
computing the same function (\cref{good-residual-ordering:fact}), and define
transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
\resleq{f}{k} ua$. This naïve approach does not yield a canonical model. For
instance, the function of \cref{non-canonical-transd:ex} can be computed in two
ways if we only consider $\resleq{f}{k}$ minimal elements, as witnessed in
\cref{non-canonical-transd:fig}. Even worse, the function $f$ belongs to
$\NSF[0]$, but one of the two presented transducers has a \kl{counter}. This is
why in \cref{residual-transducer:def}, we introduce the prefix ordering
$\prefleq$ to further constrain how the model computes.

@@

\begin{lemma}
    \label{wqo-implies-termination:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and $k \in \Nat$ be such that
    every infinite, $\prefleq$-increasing sequence is \kl{good}
    in $(\Sigma^*, \resleq{f}{k})$
    (or equivalently, such that the relation $(\prefleq \Rightarrow \resleq{f}{k})$
    is \kl{well}).
    Then, \cref{residual:algo} terminates on the input $(f,k)$.
\end{lemma}

The following \cref{n-poly-k-implies-wqo:lemma} is the core argument of the
section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
\Rightarrow \resleq{f}{k})$ is \kl{well}, but $\resleq{f}{k}$ is a
\kl{well-quasi-ordering}. This has to be compared with the property of being
\kl{$(k,\Nat)$-combinatorial}, which states that ultimately, the function is
increasing in every direction.

@@

\begin{conjecture}
    Let $f \in \ZPoly[k]$
    be \kl{combinatorial}.
    Then, $f$ is non-negative and for every pair
    $(u,w) \in \Sigma^*$,
    the sequence $\seqof{u w^n}{n \in \Nat}$
    is \kl{good}.
\end{conjecture}
\begin{proof}
    By induction on $k$.
\end{proof}

\begin{conjecture}
    Let $f \in \ZPoly[k]$
    be a non-negative function such that for every
    $(u,w) \in \Sigma^*$,
    the sequence $\seqof{u w^n}{n \in \Nat}$
    is \kl{good}.
    Then, $f$ belongs to $\NPoly[k]$.
\end{conjecture}
\begin{proof}
    By induction on $k$.
    For $k = 0$, non-negative is enough.
    To go beyond, we

\end{proof}

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ be a non-negative function, 
    and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item $f \in \NPoly[k]$;
        \item $(\Sigma^*, \resleq{f}{k})$ is a
            \kl{well-quasi-ordering};
        \item every $\prefleq$-increasing sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k}$;
        \item $f$ is \kl{computed} by a
            \kl{$k$-residual transducer}.
    \end{enumerate}
    If $f$ is asumed to be \kl{commutative}, then all of the above
    properties are decidable, and the conversions effective.
\end{theorem}
\begin{proof}
    The first two items are known to be equivalent since 
    the phd of gaëtan.
    We have proven that transducers imply
    wqo, and that wqo implies $k$-residual transducer.
\end{proof}

\subsection{Aperiodicity and Counters}
